Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-10 paper/paper.tex
Sec 5 done

\section{Experimental Evaluation}
%% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
%% we  have  built a  prototype  consisting  of  an interpreter  for  our
%% language,  a  liveness  analyzer  and  a copying  collector  that  can
%% optionally use the results of liveness analysis for marking instead of
%% reachability. When the collector uses liveness for marking, we call it
%% a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
%%   reachability-based  collector}  (RGC).   The  copying  collector  is
%% neither incremental nor generational.  As a consequence, any cell that
%% becomes unreachable or  dead is assuredly collected in  the next round
%% of garbage collection.}

@@

\subsection{copying closure using liveness vs copying closures using reachability}
To compare the  effect of using reachability to  copy closures instead
of liveness we take a couple  of programs from our benchmark suite and
execute them with both. As shown in \cred{figure-add-figure-here}, the
number of  garbage collections  in both  cases are  same but  the time
taken  varies largely.  Both these  observations can  be explained  as
follows, since we approximate the  liveness of closure to the liveness
at its  creation point, the  liveness we  get is usually  grossly over
approximated.  Using this liveness may not  give too much of a benefit
compared to  reachability as  nearly everything would  be live  at its
point of  creation.  As mentioned  in section-ref, using  liveness for
closures  may  entail  multiple  traversals over  the  same  structure
consuming more time.  Therefore we believe that using reachability for
closures is  a sweet spot in  terms of performance for  liveness based
garbage collectors.
%%
%% \comment{
%% The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
%% in the heap over time for RGC and LGC---here time is measured in terms
%% of the number of cons cells allocated. In addition, they also show the
%% number of  reachable cells and the  number of cells  that are actually
%% live\footnote{This   is  statically   approximated  by   our  liveness
%%   analysis}.  Since  the programs have  different memory requirements,
%% we have tuned the size of heap for each program to ensure a reasonable
%% number of  collections. An invocation  of RGC decreases the  number of
%% cells  in heap  till  it touches  the  curve of  reachable cells.   An
%% invocation of LGC decreases the number  of heap cells to no lower than
%% the curve of live cells.

@@

\subsection{Results}
%% \begin{figure*}[t]
%% \centering
%% \begin{boxedminipage}{.75\textwidth}
%% {\bf function} $\mathsf{copy}$($\mathsf{dfa}$, $q$, $x$)\\
%% \noindent \hspace*{.35cm}$y \leftarrow \mathsf{move\_to\_tospace}(x)$\\
%% \noindent \hspace*{.35cm}{\bf if} $x$.{\bf tag} $\neq$ \CONS\ {\bf
%%   then} {\bf skip}\\
%% \noindent \hspace*{.25cm}  {\bf else} {\bf if}
%% $\mathsf{dfa}(q,0)\mathsf{?}$ {\bf then}   $y.\CAR =
%% \mathsf{copy}(\mathsf{dfa}, \mathsf{dfa}(q,0),x.\CAR)$\\
%% \noindent \hspace*{0.95cm} {\bf if}  $\mathsf{dfa}(q, 1)\mathsf{?}$
%% {\bf then}  $y.\CDR = \mathsf{copy}(\mathsf{dfa},
%% \mathsf{dfa}(q,1),x.\CDR)$\\
%% \noindent \hspace*{.35cm}{\bf return} $y$
%% \end{boxedminipage}
%% \caption{Function       for       copying       a       root       set
%%   variable.}\label{algo:LGC-main-loop} \figrule

%% \end{figure*}

\newcommand{\hgt}{3cm}
\begin{figure*}[t]
\renewcommand{\arraystretch}{.1}
\begin{tabular}{@{}c@{}c@{}}
\hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}%, width=4.10cm}}
&
{\epsfig{file=lcss.eps, height=\hgt}}%, width=4.10cm}}
\\
\hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}%
 &
{\epsfig{file=nperm.eps, height=\hgt}}%
\\
\hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}%
&
{\epsfig{file=sudoku.eps, height=\hgt}}%%%Add knightstour.eps
\\
\hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}%
&
{\epsfig{file=treejoin.eps, height=\hgt}}%
\end{tabular}\vskip -10mm
 \caption{Memory usage  snapshot of  programs.  The  blue and  the red
   curves indicate the  number of cons cells in  the active semi-space
   for  LGC and  RGC  respectively.  The  black  curve represents  the
   number of reachable cells and  the grey curve represents the number
   of cells that are actually live  (of which liveness analysis does a
   static approximation).  x-axis is  the time  measured in  number of
   cons-cells allocated (scaled down by  factor $10^5$). y-axis is the
   number of cons-cells (scaled down by factor $10^3$).}
\label{fig:memory-usage} \figrule
\end{figure*}
The  table  in  Figure~\ref{fig:experimental-results} shows  that  LGC
improves the  garbage collection time  and efficiency for most  of the
programs.  Even for  programs where the time taken is  more it is very
competitive.  Memory usage  graphs are shown for  select benchmarks in
Figure~\ref{fig:memory-usage}.  As  the number of  garbage collections
tend  to  be  very large  our  graphs  show  only  a window  which  is
representative of the behavior for  that particular benchmark.  In all
the programs we can see that  the curve corresponding to LGC regularly
dips below the RGC curve.

One area  of concern is the  huge gap between the  actual liveness and
the liveness  percieved by  our collector.  In case  of LGC  for eager
languages  the gap  was  very  narrow and  almost  touched the  actual
liveness curve.  In case of  lazy languages, due to using reachability
for copying we  end up copying more  cells. Any data which  is part of
closure irrespective  of whether it is  live or not gets  copied. Thus
many non-live  cells get  copied during garbage  collection increasing
the drag  time.  Our  experiments have  shown that  unless we  use the
exact  execution  point  for  closures, bringing  down  the  number  of
non-live  cells  copied in  closures  cannot  be reduced.   Using  the
approximated liveness (liveness  at the creation point)  does not give
benefits  in terms  of non-live  cells inside  closures as  everything
would be live at the point of creation.
%% \comment{
%% The  increased effectiveness  of LGC  over RGC  is also  shown  in the
%% tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
%% provides  statistics  regarding the  analysis  itself.  The number  of
%% states and the analysis times  are within tolerable limits. Precision
%% of analysis refers to the percentage of dead cells that is collected by
%% LGC, averaged over all invocations.
%% The second table shows garbage  collection statistics for RGC and LGC.
%% LGC  collects larger garbage  per invocation,  drags cells  for lesser
%% time and requires a smaller heap size ({\em MinHeap}) for program to
%% run
%% in comparison with  RGC.

%% There are a couple of  issues of concern.  The garbage collection time
%% is larger in the case of LGC for some programs. The reason is that the
%% cost of consulting the liveness DFA may outweigh the combined benefits
%% of  fewer   garbage  collections   and  fewer  markings   per  garbage
%% collection.   The  other issue  is  illustrated  by  the program  {\tt
%%   lambda}    (memory   usage  shown   separately    in
%% Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
%% in Figure~\ref{fig:experimental-results}, the  number of
%% touched cells\footnote{These  are the  cells which are  visited during
%%   marking phase, often  more than once due to  sharing.} in this
%% example is much higher for
%% LGC.  This  increase is due to  excessive sharing among  heap nodes in
%% this program.  Note  that a node re-visited because  of sharing is not
%% explored  any   further  during  a  RGC   collection.   However,  this
%% curtailment cannot happen in  LGC because of the possibility that the
%% node,  re-visited in a  different liveness  state, may  mark a  set of
%% cells different from the earlier visit.
%% }
\begin{figure*}[t]
%\small
\centering
%% \scalebox{.9}{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
%% \hline
%% {Program} & {\sf sudoku} & {\sf lcss} & {\sf gc\_bench} &  {\sf
%%   nperm} &  {\sf fibheap} & {\sf knightstour} &
%% {\sf treejoin} & {\sf nqueens} & {\sf lambda}\\
%% \hline
%% \hline
%% {Time (msec)}&106.56 &2.19 &0.32 &1.16 &2.4 &3.05 &2.61 &0.71&8.68 \\
%% {DFA size} &4206 &726 &258 &526 &675 &922 &737 &241 & 694\\
%% {Precision(\%)} &98.4&98.8&99.9&87.1&100&94.3&99.6&98.8&79.7\\
%% \hline
%% \end{tabular}}
%% \centerline{(a)}
%% \medskip
%% \scalebox{0.87}{\begin{tabular}{| c | r | r | r | r | r | r  |  r | r |
%% r | r | r | r | r |}
%% \hline
%%     & \multicolumn{2}{c|}{\# Collected} & \multicolumn{2}{c|}{\#
%% Touched}
%%   & \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{} &
%% \multicolumn{2}{c|}{} & \multicolumn{2}{c|}{GC time} \\
%%                             &   \multicolumn{2}{c|}{cells per GC}
%%                             &   \multicolumn{2}{c|}{cells per GC}
%%                             &   \multicolumn{2}{c|}{\#GCs}
%%                             &   \multicolumn{2}{c|}{MinHeap}
%%                             &   \multicolumn{2}{c|}{Avg. Drag}
%%                             &   \multicolumn{2}{c|}{(sec)} \\
%% \cline{2-13}
%% {Program}    &
%% RGC & LGC & RGC & LGC  & RGC & LGC  &   RGC & LGC & RGC & LGC & RGC &
%% LGC \\
%% \hline
%% \hline
%%     {\sf   sudoku}  &381 &1107 &1268 &281 &169 &159 & 1346  &338 &527 &5
%% &.008 & .026 \\
%%     {\sf  lcss } & 46522 &51101 &6216 &1363&30&24& 52301  &1701 &5147
%% &588 &.007 & .008 \\
%%     {\sf   gc\_bench}  &129179 &131067 &1894 &4&34&4& 131071   &6 &16970
%%  &4 &.13 & .0002 \\
%%      {\sf  nperm}  & 47586  &174478 &201585 &60882&710&386& 202597  &37507
%% &171878 &76618 &0.250 & .154  \\
%%     {\sf  fibheap} &249502  &251525 &5555 &2997&1002&975& 254520  &13558
%% &78720 &0 &1.185 & 5.457  \\
%%     {\sf  knightstour}  &2593 &314564 &907502 &319299&1161&10&508225
%% &307092 &206729 &82112 &464.902 & 14.124  \\
%%     {\sf  treejoin} & 288666  &519943 &297570 &5547&116&81& 525488  &7150
%% &212653 &1954 &4.335 & 4.4 \\
%%     {\sf   nqueens} & 283822 &1423226 &2133001 &584143&46&9& 1819579
%% &501093 &521826 &39465 &70.314 & 24.811 \\
%%     {\sf   lambda}  &205 & 556 &2072&90345 &775 &750&966 & 721  &303 &95
%% &.179 &1.01  \\
%% %    {\sf   fft }& &  & & &&&& ???? &???? &????? &????? &???? & ????
%% \\
%% %    {\sf  nboyer} & &  & &   & &  & & &  \\
%% %    {\sf   circsim?} & &  & &   & & & & &   \\
%%  \hline
%% \end{tabular}}
%% \centerline{(b)}

\include{autotable}
\vskip -0.75mm
\caption{Experimental results comparing RGC and LGC. Table (a) gives
  data related to liveness analysis, and (b) gives garbage collection
  data. The average drag time is measured in allocated cons cells.}
\label{fig:experimental-results}
\normalsize
\end{figure*}

%% \section{Collecting more garbage can never slow things down}
%% \label{sec:lgc-always-better}

%% Since garbage collection is  effectively asynchronous to the
%% allocator
%% thread, one might  worry as to how {\em  robust} our measurements
%% are.
%% For example,  while LGC would,  in general, collect more  garbage
%% than
%% RGC in the same heap state,  is it possible that on some programs,
%% LGC
%% might do  a larger number  of collections? We  prove a result  to
%% show
%% that   this  cannot   happen.    This  result   applies  to
%% classical
%% mark-and-sweep and copying garbage  collectors and, we hypothesize,
%% to
%% generational collectors too.

%% \begin{lemma}
%% For the  same mutator,  a liveness-based collector  can never  do
%% more
%% garbage collections than a reachability-based collector.
%% \end{lemma}

%% \begin{proof}Assume, as before,  that time is measured in terms of
%% the number of
%%   cons cells allocated. Let us run  two copies of the mutator with
%% the
%%   two    different   collectors    but    with   synchronous
%% memory
%%   allocations.   The   garbage    collections,   however,   happen
%%   asynchronously.

%%   To prove the lemma, it is enough to show the truth of the
%% following
%% statement: After
%%  every  LGC invocation,  the count of LGC invocations   is  no
%% greater
%% than  RGC invocations.   The  base case  holds  since the  first
%% invocations of both GCs happen at the same time.  Assume the
%% statement
%% to be true after $n$ invocations of LGC.  Since LGC copies a subset
%% of
%% reachable cells, its heap would contain no more cells than RGC heap
%% at
%% the end  of the $n$th invocation.   Thus either RGC  is invoked next
%% before
%% LGC, or LGC and RGC are both invoked next at the same time.  In
%% either
%% case, the statement holds after $n+1$ invocations of LGC.
%% \end{proof}

%% \begin{figure}[h]
%% \centerline{\epsfig{file=memory-usage.eps, width=12cm, height=15cm}}
%% \caption{Memory  usage of  programs. The  solid  lines represents
%%   number of  live cells and the  dashed lines show  the number of
%%   reachable cells.  The red and blue  lines represent invocations
%%   of RGC and  LGC---their heights indicate  the amount of memory
%%   copied.   {\color{red}{Caution:  Figures  are from  a  previous
%%       paper and are being used as placeholder}}}
%% \label{fig:memory-usage} \figrule
%% \end{figure}
%%\clearpage

>>>>

\begin{table*}[t!]
\caption{Experimental results comparing RGC and LGC}
\label{tab:experimental-results}

\centering
\include{autotable}
\end{table*}
\section{Experimental Evaluation}

%% \comment{To demonstrate the effectiveness of liveness-based garbage collection,
%% we  have  built a  prototype  consisting  of  an interpreter  for  our
%% language,  a  liveness  analyzer  and  a copying  collector  that  can
%% optionally use the results of liveness analysis for marking instead of
%% reachability. When the collector uses liveness for marking, we call it
%% a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
%%   reachability-based  collector}  (RGC).   The  copying  collector  is
%% neither incremental nor generational.  As a consequence, any cell that
%% becomes unreachable or  dead is assuredly collected in  the next round
%% of garbage collection.}

@@

\subsection{copying closure using liveness vs copying closures using reachability}
To compare the  effect of using reachability to  copy closures instead
of liveness we take a couple  of programs from our benchmark suite and
execute them with both. As shown in \cred{figure-add-figure-here}, the
number of  garbage collections  in both  cases are  same but  the time
taken  varies largely.  Both these  observations can  be explained  as
follows, since we approximate the  liveness of closure to the liveness
at its  creation point, the  liveness we  get is usually  grossly over
approximated.  Using this liveness may not  give too much of a benefit
compared to  reachability as  nearly everything would  be live  at its
point of  creation.  As mentioned  in Section~\ref{sec:live-clo}, using  liveness for
closures  may  entail  multiple  traversals over  the  same  structure
consuming more time.  Therefore we believe that using reachability for
closures is  a sweet spot in  terms of performance for  liveness based
garbage collectors.
%%
%% \comment{
%% The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
%% in the heap over time for RGC and LGC---here time is measured in terms
%% of the number of cons cells allocated. In addition, they also show the
%% number of  reachable cells and the  number of cells  that are actually
%% live\footnote{This   is  statically   approximated  by   our  liveness
%%   analysis}.  Since  the programs have  different memory requirements,
%% we have tuned the size of heap for each program to ensure a reasonable
%% number of  collections. An invocation  of RGC decreases the  number of
%% cells  in heap  till  it touches  the  curve of  reachable cells.   An
%% invocation of LGC decreases the number  of heap cells to no lower than
%% the curve of live cells.

@@

\subsection{Results}
\newcommand{\hgt}{3cm}
\begin{figure*}[t!]
\renewcommand{\arraystretch}{.1}
\begin{tabular}{@{}c@{}c@{}}
\hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}%, width=4.10cm}}
&
{\epsfig{file=lcss.eps, height=\hgt}}%, width=4.10cm}}
\\
\hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}%
 &
{\epsfig{file=nperm.eps, height=\hgt}}%
\\
\hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}%
&
{\epsfig{file=sudoku.eps, height=\hgt}}%%%Add knightstour.eps
\\
\hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}%
&
{\epsfig{file=treejoin.eps, height=\hgt}}%
\end{tabular}\vskip -10mm
 \caption{Memory usage  snapshot of  programs.  The  blue and  the red
   curves indicate the  number of cons cells in  the active semi-space
   for  LGC and  RGC  respectively.  The  black  curve represents  the
   number of reachable cells and  the grey curve represents the number
   of cells that are actually live  (of which liveness analysis does a
   static approximation).  x-axis is  the time  measured in  number of
   cons-cells allocated (scaled down by  factor $10^5$). y-axis is the
   number of cons-cells (scaled down by factor $10^3$).}
\label{fig:memory-usage} \figrule
\end{figure*}

Table~\ref{tab:experimental-results} shows  that  LGC
improves the  garbage collection time  and efficiency for most  of the
programs.  Even for  programs where the time taken is  more it is very
competitive.  Memory usage  graphs are shown for  select benchmarks in
Figure~\ref{fig:memory-usage}.  As  the number of  garbage collections
tend  to  be  very large  our  graphs  show  only  a window  which  is
representative of the behavior for  that particular benchmark.  In all
the programs we can see that  the curve corresponding to LGC regularly
dips below the RGC curve.

One area  of concern is the  huge gap between the  actual liveness and
the liveness  percieved by  our collector.  In case  of LGC  for eager
languages  the gap  was  very  narrow and  almost  touched the  actual
liveness curve.  In case of  lazy languages, due to using reachability
for copying we  end up copying more  cells. Any data which  is part of
closure irrespective  of whether it is  live or not gets  copied. Thus
many non-live  cells get  copied during garbage  collection increasing
the drag  time.  Our  experiments have  shown that  unless we  use the
exact  execution  point  for  closures, bringing  down  the  number  of
non-live  cells  copied in  closures  cannot  be reduced.   Using  the
approximated liveness (liveness  at the creation point)  does not give
benefits  in terms  of non-live  cells inside  closures as  everything
would be live at the point of creation.
